Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-10 latex/main.tex
UPdates

The genomic revolution has brought about a paradigm shift
in the way we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER\cite{Turakhia2021,DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

@@

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model\cite{LiStephens2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{sc2ts_methodology}
    \caption{Schematic overview of sc2ts. (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. Sample 0 is used to initialize the ARG as the root. Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child. The resulting ARG contains two local trees, one on each side of the inferred breakpoint. (B) Copying paths for Sample 4 inferred under the LS model while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).}
    \label{fig:methodology}
\end{figure}

@@

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
from the Viridian UShER phylogeny\cite{Hunt2024} and a previous study\cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in
the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango Xs are XA,
which is the first designated Pango X that showed signs of onward transmission\cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence\cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the 
ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango Xs were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with the XA.
Two "Group A" recombinant samples reported by Jackson et al.\cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach:
they scan candidate recombinant sequences for mutations associated with different Pango lineages
without using a phylogeny.

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table \ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies\cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021} (Table \ref{tab:jackson}).

@@

[FIXME - discuss BA.5 and the BA.2 event instead.]
Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors.
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling if still possible.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
\label{fig:averted_mutations}
\end{figure}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous 
studies\cite{Turakhia2022,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{Recombination intervals. (A) Distribution of the breakpoint
intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top). Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
\label{fig:recombinant_intervals}
\end{figure}

>>>>

The genomic revolution has transformed how we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

@@

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{sc2ts_methodology}
\caption{
Schematic overview of sc2ts.
(A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates.
Sample 0 is used to initialize the ARG as the root.
Next, Sample 1 is attached, mapping mutations onto the new branch to explain it.
When Sample 4 is detected as a recombinant,
a node representing the recombination event is inserted with two edges connecting to its parents,
with Sample 4 itself attached to this node as its child.
The resulting ARG contains two local trees,
one on each side of the inferred breakpoint.
(B) Copying paths for Sample 4 inferred under the LS model
while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).
}
\label{fig:methodology}
\end{figure}

@@

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions
in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ''X'', e.g., XA and XBB).
Below, we summarize our analysis of the Pango X lineages in the ARG
(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass the robustness filtering criteria (Figure~\ref{fig:recombinant_qc}).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with the XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
RecombinHunt \cite{Alfonsi2024}, CovRecomb \cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach:
they scan candidate recombinant sequences for mutations 
associated with different Pango lineages
without using a phylogeny.

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb,
we summarized the detection results for these Pango X lineages (if available)
reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}).

@@

[FIXME - discuss BA.5 and the BA.2 event instead.]
Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors.
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling, if possible.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
\label{fig:averted_mutations}
\end{figure}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous studies \cite{Turakhia2022,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination intervals.
(A) Distribution of the breakpoint intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
\label{fig:recombinant_intervals}
\end{figure}
